---
title: Competing risks and the Fine-Gray model
author: Terry Therneau
date: March 2023
output: beamer_presentation
---

<!-- A talk for the division, March 2023 -->

```{r, echo=FALSE}
library(survival)
library(splines)
knitr::opts_chunk$set(comment=NA, tidy=FALSE, echo=FALSE,
                      fig.height=6, fig.width=8,
                      warning=FALSE, error=FALSE)
options(contrasts= c("contr.treatment", "contr.poly"),
        show.signif.stars = FALSE, continue = " ")
par(mar=c(5,5,1,1))
table2 <- function(...) table(..., useNA= "ifany")

#palette("Okabe-Ito")
data2 <- readRDS("data2.rds")  # from the long-term follow-up study
data2$year1 <- 1*(data2$year==0)
# collapse, for simpler presentation
d2 <- survcondense(Surv(age1, age2, state) ~ male + educ + apoepos + icmc + 
                          cstate + year1, data2, id= clinic)
d2$edu4 <- pmax(d2$educ, 8)/4
d2$cmc2 <- d2$icmc/2
d2$cmc  <- d2$icmc

# The contrast function is taken from 
#      adir/Therneau/longterm.followup/tmt/marginal.Rnw
#
contrast <- function(fit, data, global=FALSE, weight, 
                     tol=sqrt(.Machine$double.eps)) {
    newx <- model.matrix(fit, data= data)
    if (global)  # test for all differences =0
        newx <- scale(newx[-1,], center=newx[1,], scale=FALSE)
    else {    
        if (missing(weight))  # subtract first row
            newx <- scale(newx, center=newx[1,], scale=FALSE) 
        else  { # use a weighted average
            wt <- as.vector(weight/sum(weight)) # use a weighted average
            if (length(wt) != nrow(newx)) stop("wrong length for weight")
            newx <- scale(newx, center= wt %*% newx , scale=FALSE)
        }
    }
    test <- drop(newx %*% coef(fit))
    V    <- newx %*% vcov(fit) %*% t(newx)
    if (global) {
        stemp <- svd(V)
        nonzero <- (stemp$d > tol)
        ctemp <- test %*% stemp$u[,nonzero]
        chi <- ctemp %*% diag(1/stemp$d[nonzero]) %*% c(ctemp)
        c(chisq= drop(chi), df= sum(nonzero))
    }
    else {
        std <- sqrt(diag(V))
        z <- ifelse(std==0, 0, test/std)
        cbind(estimate=test, std.err=std, z= z)
    }
}
```
## Free light chain
```{r, c1}
fdata <- subset(flchain, futime > 7)  # drop the eary deaths
fdata$id <- 1:nrow(fdata)
fdata$years <- fdata$futime/365.25
fdata$flc10 <- 1*(fdata$flc.grp ==10)

temp <-  with(fdata, ifelse(death==0, -1, 
                            1*(chapter=="Circulatory") +
                            2*(chapter=="Neoplasms") +
                            3*(chapter=="Respiratory")))
fdata$state <- factor(temp, c(-1, 1,2, 3, 0), 
                      c("censor", "CVD", "Cancer", "Resp", "Other"))

temp <- table2(fdata$state)

sname <- paste(c("Entry", "CVD", "Cancer", "Resp", "Other"),
               c(nrow(fdata), temp[2:5]), sep='\n')
smat <- matrix(0, 5,5, dimnames=list(sname, sname))
smat[1, 2:5] <- 1
statefig(c(1,4), smat)
```

---

```{r, fanal, echo = TRUE}
# create a factor (class) variable (hidden)
table(fdata$state)

# Aalen-Johansen
fsurv <- survfit(Surv(years, state) ~1, data=fdata, id=id)

# Multi-state hazard model
fcox  <- coxph(Surv(years, state) ~ age + sex + flc10, 
               data= fdata, id = id)
```

---

```{r, ag1}
oldpar <- par(mfrow=c(1,2), mar=c(5,5,1,1))
plot(fsurv, col=1:5, lwd=2, noplot="", lty=c(1,1:4),
     xlab="Years", ylab="P(state)")
legend(0, .7, c("Alive", "CVD", "Cancer", "Resp", "Other"), lwd=2,
       col=1:5, lty=c(1, 1:4), bty='n')
plot(fsurv, col=2:5, lwd=2, xscale=365.25, lty=c(1:4),
     xlab="Years", ylab="P(state)")
par(oldpar)

```

---

The coxph call produces a multi-state hazard model fit.  One set of
coefficients for each transition (arrow) in the diagram

```{r, coef}
ctemp <- coef(fcox)
stemp <- sqrt(diag(vcov(fcox)))
xtemp <- paste0(round(exp(ctemp),2), "(", round(ctemp/stemp,1), ")")
xtemp <- matrix(xtemp, ncol=4)
dimnames(xtemp) <- list(c("age", "male", "FLC"), 
                        c("CVD", "Cancer", "Resp", "Other"))
print(xtemp, quote=FALSE)
```

---

## What is the effect of FLC on *cancer* death?
Create predicted curves based on the fitted model.

```{r, cvdeath, echo=TRUE}
dummy <- expand.grid(flc10 = 0:1,  sex = c("F", "M"), 
                     age= c(50,60, 70, 80))
predsurv <- survfit(fcox, newdata=dummy)
dim(predsurv)
```

---

```{r, tplot}
yr <- c(0, .18)
oldpar <- par(mfrow=c(2,2), mar=c(5,5,1,1))
plot(predsurv[1:4, 3], lty=2:1, col=c(1,1,2,2), lwd=2, ylim=yr,
	xlab="Years", ylab="P(cancer)")
legend(0, .17, c("M, FLC high", "F, FLC high", "M, FLC low", "F, FLC low"),
       lty=c(1,1,2,2), col=2:1, lwd=2, bty='n', cex=.9)
text(10, .01, "Age 50")
plot(predsurv[5:8, 3], lty=2:1, col=c(1,1,2,2), lwd=2, ylim=yr,
	xlab="Years", ylab="P(cancer)")
text(10, .01, "Age 60")
plot(predsurv[9:12, 3], lty=2:1, col=c(1,1,2,2), lwd=2, ylim= yr,
	xlab="Years", ylab="P(cancer)")
text(10, .01, "Age 70")
plot(predsurv[13:16, 3], lty=2:1, col=c(1,1,2,2), lwd=2, ylim= yr,
	xlab="Years", ylab="P(cancer)")
text(10, .01, "Age 80")
par(oldpar)

```

---

```{r, tsum}
stemp <- summary(predsurv, time=c(7, 14))$pstate[,,3]  # pick off Cancer
# I want to present this as a table
temp2 <- array(stemp, dim=c(2,2,8))
delta <- temp2[,2,] - temp2[,1,]
temp3 <- t(delta * 100)
dimnames(temp3) <- list(paste(rep(paste("Age", c(50,60,70,80)),each=2),
                              rep(c("Female", "Male"),4), sep=', '),
                        c("7 year", "14 year"))
round(temp3,2)
```
But the investigator wants a 1 number summary.

---

## Additive models
- The three most popular models in statistics
   * Linear: $E(y) = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \ldots$
   * GLM: $E(y) = g\left(\beta_0 + \beta_1 x_1 + \beta_2 x_2 + \ldots \right)$
   * Cox: $\lambda(t) = \exp\left(\beta_0(t) + \beta_1 x_1 + \beta_2 x_2 + \ldots \right)$
- Why?  Simplicity.
   * If `x1= FLC+`, then $\beta_1$ is *THE* effect of FLC, independent of any other variables in the model.
   * Statisticians like this.
   * Investigators really like this (a single p-value)
- (Generalized additive models will replace one of the $\beta x$ terms
      with $s(x)$, but retain the separability.)

## 3 criteria for a successful statistical model
1. Simplicity: in the sense described above, leading to simple explanations
    for the effect of key predictors.
2. Statistical validity: the model must describe the data adequately. ``All 
models are wrong.  The practical question is whether a model is wrong enough to
not be useful.'' George Box
3. Numerical stability: the code to fit a model does not require hand-holding
or fiddling with tuning parameters: it just runs.

The transform $g$ gets chosen to fit criteria 3; if it helps with criteria 2 
that is mostly luck.  (It nearly always impedes interpretability).

## Fine-Gray: key idea
For an ordinary 2 state Cox model:
$$ 
	P(death) = g(\beta_0(t) + \beta_1 x_1 + \beta_2 x_2 + \ldots)
$$
where $g$ = complementary log-log

Assume that for outcome $k$
$$
 p_k(t)= g(\beta_{k0}(t) + \beta_{k1} x_1 + \beta_{k2} x_2 + \ldots)
$$

Issues
- *how* to fit this (censored data)
- is it a sensible model?


## Transforms
```{r, logit}
x <- seq(-4, 4, length=101)
logit <- function(x) exp(x)/(1 + exp(x))
cloglog <- function(x) exp(-exp(1-x))
matplot(x, cbind(logit(1.2*x), pnorm(x), cloglog(x+1.3)), type='l', 
        lty=3:1, lwd=3, xlab="Risk score", ylab="P(fail)")
legend(-4, .9, c("Logistic", "Probit", "Clog-log"), lty=3:1, lwd=2, 
       col=1:3, bty='n')
```

## Example
```{r, states}
oldpar <- par(mar=c(1,1,1,1))
cstate <- c("CU", "dementia", "death")
cmat <- matrix(0, 3,3, dimnames=list(cstate, cstate))
cmat[-3,3] <- cmat[1,2] <- 1
statefig(c(1,2), cmat, cex=1.2)
par(oldpar)
```


## Predictors of dementia and death
- `r length(unique(data2$ptnum))` subjects
- 726 dementia, 1990 deaths, 1/2 the dementias occur after active participation
- Taken from the MCSA, an age/sex stratified random sample from Olmsted County, Minnesota
- REP infrastructure
- Covariates
    * APOE e4 allele: risk factor for amyloidosis
    * CMC score: 0-7, count of morbidities
	
---

```{r, rateplot}
d3 <- subset(d2, cstate=="ND")
oyear <- with(d3,  age2-age1)   # interval length
death0 <- glm((state=='death') ~ age1 + male + year1 +offset(log(oyear)), 
             poisson, data=d3)
death1 <-glm((state=='death') ~ ns(age1,3) + male + year1 +offset(log(oyear)), 
             poisson, data=d3)
death2 <- update(death1, . ~ . + male:age1)

dement1 <- glm((state=="dementia") ~ ns(age1,3) + male +
                     offset(log(oyear)), poisson, d3)
dement2 <- glm((state=="dementia") ~ ns(age1,3)+ male + age1:male+  
                     offset(log(oyear)), poisson, data=d3)

# anova(death0, death1, death2, test="Chisq")
# anova(dement1, dement2, test="Chisq")

# no evidence for a sex by rate interaction

dummy1 <- expand.grid(age1=60:95, male=0:1, year1=0, cmc=1, oyear=100)
yhat1 <- matrix(predict(death0, newdata=dummy1, type='response'), ncol=2)
yhat2 <- matrix(predict(dement1, newdata=dummy1, type='response'), ncol=2)

oldpar <- par(mar=c(5,5,1,1))
matplot(60:95, cbind(yhat1,yhat2), type='l', lty=c(2,2,1,1), col=1:2, log='y',
        lwd=2, xlab="Age", ylab= "Events per per 100")
legend("topleft", c("M death", "F death", "M dementia", "F dementia"),
       col=2:1, lty=c(2,2,1,1), bty='n', lwd=2)
#matlines(60:95, .95*survexp.mn[61:96,,"2013"]*36525, col="gray70", lty=3)

par(oldpar)
```

## Multistate
```{r}
oldpar <- par(mar=c(1,1,1,1))
statefig(c(1,2), cmat, cex=1.2)
par(oldpar)
```

---

```{r, coxfit0}
cfit <-  coxph(list(Surv(age1, age2, state) ~ apoepos + male + cmc2 + edu4,
                    1:2 ~ apoepos:male + year1),
                    d2, id=clinic, istate=cstate)
cfit0 <- coxph(Surv(age1, age2, state) ~ apoepos + male + cmc2 + edu4,
                    d2, id=clinic, istate=cstate)
temp <- summary(cfit0)$coef
temp2 <- temp[,c(1,3)]
temp3 <- exp(temp2[,1] + outer(temp2[,2], c(0, -1.96, 1.96), "*"))
oldpar <- par(mar=c(5, 8, 1, 1))
yy <- c(14:11, 9:6, 4:1)
matplot(temp3, yy, log='x', type='n', xlab="Hazard ratio", 
       ylab="", yaxt='n')

points(temp3[,1], yy, pch=rep(15:17, c(4,4,4)), col=rep(1:3, c(4,4,4)))
segments(temp3[,2], yy, temp3[,3], yy, col=rep(1:3,c(4,4,4)))
abline(h=c(5, 10), lty=2)
abline(v=1, lty=2)
ylab <- rep(c("APOE+", "Male", "CMC(2)", "Educ(4)"), 3)
axis(2, yy, ylab, las=2)
text(c(1.7, 1.7, 1.6), c(1.5, 6.1, 11), 
     c("dementia : death", "CU : death", "CU : dementia"))
par(oldpar)
```

## Competing risks
```{r, doublefig}
oldpar <- par(mfrow=c(1,2), mar=c(1,1,1,1))
statefig(1:2, cmat, cex=1.2)
cmat2 <- cmat
cmat2[2,3] <- 0
temp<- c("CU","dementia\n before death", "death before\ndementia")
dimnames(cmat2) <- list(temp, temp)
statefig(1:2,cmat2)
par(oldpar)
```

---

```{r, coxfit2}
# for a competing risks plot, don't let them progress
data3 <- subset(d2, cstate== "ND")
cfit0b <- coxph(Surv(age1, age2, state) ~ apoepos + male + cmc + edu4,
                   data3, id=clinic)

dummy <- expand.grid(male =0:1, apoepos =0:1, cmc= c(1,3), edu4=c(3, 4))
surv2 <- survfit(cfit0b, newdata= dummy, start.time=60, p0=c(1,0,0))
oldpar <- par(mfrow=c(2,2), mar=c(5,5,1,1))
plot(surv2[1:4,2], col=1:2, lty=c(2,2,1,1), lwd=2, xmax=100, 
     xlab="Age", ylab="P(Ever dementia)")
legend(60, .45, c("Female+", "Male+", "Female-", "Male-"), col=1:2, 
       lty=c(1,1,2,2), lwd=2, bty='n')
text(95, .05, "CMC=1\neduc=12")
plot(surv2[5:8,2], col=1:2, lty=c(2,2,1,1), lwd=2, xmax=100, 
     xlab="Age", ylab="P(Ever dementia)")
text(95, .05, "CMC=3\neduc=12")

plot(surv2[1:4,3], col=c(1,2,1,2), lty=c(2,2,1,1), lwd=2, xmax=100,
     xlab="Age", ylab="P(death w/o dementia)", ylim=c(0, .7))
legend(60, .55, c("Female+", "Male+", "Female-", "Male-"), col=1:2, 
       lty=c(1,1,2,2), lwd=2, bty='n')
text(95, .05, "CMC=1\nEduc=12")
plot(surv2[5:8,3], col=1:2, lty=c(2,2,1,1), 
     lwd=2, xmax=100, xlab="Age", ylab="P(death w/o dementia)")
text(95, .05, "CMC=3\nEduc=12")
par(oldpar)
```

## Fine-Gray
- The effect of sex on P(dementia) depends on the levels of all the other 
covariates, and on time.
- There is no single p-value. 
-
- Model the two outcomes directly:
   * P(dementia before death) = $g(\beta_0(t) + X\beta)$ 
   * P(death before dementia) = $g(\alpha_0(t) + X\alpha)$
   * $g$ = the complimentary log log
- Technical challenge.
   * Treating survival as binomial


## Geskus' formulation
> - Create a special data set for each outcome.
    * Subjects who are censored persist, but with diminished case weights.
    * Weights decrease based on $F(t)$ and $G(t)$.
> - (You can't have time-dependent covariates.)
> - Apply an ordinary Cox model program to the new data set.
- Advantage: all the Cox model checks are available.
> - For the dementia dataset, subjects who die also persist, but with diminished
case weights.

## Geskus
```{r, geskus, echo=TRUE}
fdata1 <- finegray(Surv(age1, age2, state) ~., data=data3, 
                   id= clinic, etype= "dementia")
fdata2 <- finegray(Surv(age1, age2, state) ~., data=data3, 
                   id= clinic, etype= "death")
#
rbind(data3 = dim(data3), fdata1 = dim(fdata1), fdata2= dim(fdata2))

fcox1 <- coxph(Surv(fgstart, fgstop, fgstatus) ~ apoepos + 
                male + cmc + edu4, weight = fgwt, fdata1)
fcox2 <- coxph(Surv(fgstart, fgstop, fgstatus) ~ apoepos + 
                male + cmc + edu4, weight = fgwt, fdata2)
```

---

```{r, geskus2}
temp <- rbind("multi, CU:dementia" = coef(cfit0)[1:4],
               "multi, CU:death" = coef(cfit0)[5:8],
              "FG, dementia before death" = coef(fcox1), 
              "FG, death before dementia" = coef(fcox2))
colnames(temp) <- names(coef(fcox2))
round(temp,3)

cat("cox.zph on fcox1\n")
zp <- cox.zph(fcox1, transform= "identity")
zp
``` 

## How well does it work?
- If cause 2 has low prevalence ($<1/4$) and/or cause 2 has no strong
covariates, then all is well for the cause 1 model
   * Coefficients for outcome 1 hardly change from the Cox model
   * The predicted curves have the same shape, but are attenuated
- Examples
   * Revision after hip fracture
   * Epidemic

## Otherwise
1. The model often does not fit very well.  It fails our 'good enough' criteria.
2. There is no physical system that satisfies the FG model.
3. Users interpret coefficients as though it were a Cox model, and it is not.
4. It encourages bad science: most examples (and users) fit only one of the endpoints, ignoring the other.
5. If there are moderately strong covariates, and $>80$\% reach 
one of the two endpoints, it is common to have $\hat P$(dementia before death) + $\hat P$(death before dementia) $> 1$
for high risk subjects.
6. In the guts of the code, people who die are still at risk for
dementia.

## What to do?
1. Intentionally report both hazards and absolute risk
   - Biology + the consequences of that biology
   - A two number summary of (APOE hazard, APOE risk), true for all time, 
for all combinations of other covariates, is an impossible dream.
   - For absolute risk, choose 1 (or 2) timepoints of interest.
   - Use pseudovalues or marginal estimates for those time points.
2. Marginal estimates
   - If APOE is the variable of interest, average over the others
       * dummy data set with $n$ rows, everyone APOE-
       * get all $n$ predicted curves, take the average
       * repeat for APOE+
   - g-estimation
3. Pseudovalues
   - From the appropriate KM or Aalen-Johansen (CI) curve
   - Select one or more time points, and create the matrix of *pseudovalues*
   - Essentially, the influence of each observation on $p(t)$
   - Use these in a regression model

---

```{r, pseudo, echo= TRUE}
ajfit <- survfit(Surv(age1, age2, state) ~ 1, id = clinic, 
                 data=data3, start.time = 65)
pdat <- pseudo(ajfit, times= c(70, 80, 90, 100))
dim(pdat)
d100 <- pdat[,4,2]  # influence on dementia at age 100

# data with one obs per subject
base <- subset(data3, !duplicated(clinic))
pfit1 <- glm(d100 ~ apoepos + male + cmc +edu4, base,
            family= gaussian(link = blogit()))
pfit2 <- glm(d100 ~ apoepos + male + cmc +edu4, base,
            family= gaussian(link = bcloglog()))
```

---

```{r, psplot}
plot(ajfit, noplot="", col=c(1,3,2), lwd=2,  xmax=100,
     ylab="P(state)")
text(c(70, 95, 98), c(.88, .68, .37), c("CU", "death", "dementia"), 
     col=1:3)
```

---

```{r, pseudo2}
temp <- rbind("pseudo, logit" = pfit1$coef, 
              "pseudo, cloglog" = pfit2$coef,
              "FineGray, dem" = c(NA, coef(fcox1)),
              "multistate HR"  = c(NA, cfit0$coef[1:4]))
print(round(temp,2), na.print="")
```

## Multiple time points
- For multiple time points at once: 
    * A bit more work to set up.
    * Add `factor(time)` to the fit: one intercept per time point.
    * Robust variance is necessary, fit using GEE instead of glm.
- Closely related to ordinal logistic regression
- With many time points, result will approach the FG 
   * Coefficients will be nearly identical, se a small bit larger
   * Adding time*covariate interactions is a test for 'proportional cloglog'
   * A good way to more deeply understand the Fine Gray model


## Final
- Multi-state models are important
    * No one outcome is dominant
    * Want to understand the trajectory of disease
    * Both rates and outcomes are necessary summaries
- We like additive models.
- Additive on hazard scale $\ne$ additive on absolute risk scale
- FG was an early attempt to address this.  Credible at the time, but has
not aged well.
- It works when you don't need it, and fails when you do.

---

```{r, lastplot}
dummy0 <- expand.grid(apoepos=0:1, male=0:1, cmc2= 1, cmc=2, edu4= 4)
csurv <- survfit(cfit0, newdata=dummy0, start.time=65)

plot(csurv[,2], col=c(1,1,2,2), lty=1:2, lwd=2, xmax=100,
     xlab="Age", ylab="P(dementia)")
legend(65, .1, c("Female", "Male"), col=1:2, lwd=2, bty='n')
text(c(80, 89), c(.08, .035), c("APOE+", "APOE-"))
```

---

```{r, last2}
surv2a <- survfit(cfit0b, newdata=dummy0, start.time=65, p0=c(1,0,0))
surv2b <- survfit(cfit0b, newdata=dummy0, start.time=80, p0=c(1,0,0))
plot(surv2a[1:4,2], lty=1:2, col=c(1,1,2,2), lwd=2, xlab="Age",
     ylab= "Dementia before death", xmax=100)
lines(surv2b[1:4, 2], col=c(3,4,3,4), lwd=2, lty=c(1,1,2,2))
legend(75, .4, c("Male +", "Female +", "Male -", "Female -"),
       col=c(1,2,1,2), lty=c(2,2,1,1), bty='n', lwd=2)
```
