---
title: Suvival workshop, part 4
author: Terry Therneau
output: pdf_document
---

```{r, setup, echo=FALSE}
library(knitr)
knitr::opts_chunk$set(comment=NA, tidy=FALSE, highlight=FALSE, echo=FALSE,
               warning=FALSE, error=FALSE)
library(survival)
library(splines)
```
# MCSA
I have learned a tremendous amount about multi-state models from an analysis
of the Mayo Clinic Study of Ageing (MCSA) data.
One paper's analysis in particular (doi: 10.1093/braincomms/fcac017) has a
directory with 14 different 'exporatory' knitr files.

```{r, anal}
load("data/mcsa.rda")   # this data is not shared
survcheck(Surv(age1, age2, state) ~ 1, data= mcsa, id= ptnum, istate= cstate)

cfit1  <- coxph(list(Surv(age1, age2, state) ~ iclgp + apoepos + male +
                         educ4 + icmc, 
                    1:2 ~ iclgp:male + apoepos:male, 1:3 ~ yearc),
                    id=ptnum, mcsa, istate= cstate)

print(cfit1, digits=2)
```

```{r, fig.align='center', out.width='0.9\\linewidth'}
include_graphics("figures/efig2.pdf")
```

```{r, fig.align='center', out.width='0.9\\linewidth'}
include_graphics("figures/efig4.pdf")
```

```{r, fig.align='center', out.width='0.9\\linewidth'}
include_graphics("figures/efig3.pdf")
```

Males have higher hazard of dementia,  equal lifetime risk, smaller sojourn 
years.

There are a lot of choices in this fit.
* The amlyoid level is treated as categorical, to deal with missing.
* An APOE by sex interaction is known from the literature, we sort of have to
  add it.  The amyloid by sex interaction was also significant.
* Age scale makes sense
* There is an enrollment effect on the CU:death transition. But it doesn't
work for the CU:dementia one, due to the 15 month visit interval.  We don't
expect an enrollment time effect on dementia:death.
* This fit used initial amyloid level and intial CMC, which I later realized
was worrisome.

# Time dependent covariates and age scale
Some rules that I have stated over the years
* TD covariates are very useful in Cox models
* Survival curves + time dependent covariates doesn't make sense
* Use landmark curves
* Age scale is superior to entry scale for many problems
* Multi-state models are preferred
* In a mulit-state model, both HR and absolute risk are needed

For the MCSA paper we have time dependent covariates that *will* rise for
a lot of people (amyloid and CMC), age scale makes the most sense, absolute
risk is critical.  But: the landmark approach no longer makes sense.

```{r, amyloid}
alldat <- readRDS("data/MCSA-all-visits.rds")
mdata <- subset(alldat, !is.na(pzmemory) & agevis > 50 & !is.na(apoepos), 
                c(clinic, agevis, cyclenum, date, male, educ, apoepos, pzmemory,
                  pzglobal, spm12.pib.ratio, spm12.tau.ratio, 
                  pibdate, taudate,testnaive))
temp <- c("agevis", "spm12.pib.ratio", "spm12.tau.ratio")
names(mdata)[match(temp, names(mdata))] <- c("age", "pib", "tau")

ptemp <- subset(mdata, !is.na(pib))
acount <- table(ptemp$clinic)
# use a subset of all those with 4 or more
id4 <- names(acount)[acount >3]
temp4 <- subset(ptemp, clinic %in% id4)

# spread out over age
temp <- id4[order(temp4$age[!duplicated(temp4$clinic)])]
id4b <- temp[seq(1, length(temp), length=60)]

plot(pib ~ age, temp4, subset=(clinic %in% id4b),
     log='y', type='n',  xlab="Age", ylab="Amyloid")
for (i in 1:length(id4b)) {
    lines(pib ~ age, temp4, subset= (clinic== id4b[i]), lwd=2,
         col= 1 + i%%5, lty= 1)
}
```

Problem:
* consider the risk set for a dementia at age 82.35
* amyloid-at-baseline as a covariate
* for some subjects that value is 1 month old, for some 10 years old.
* what does the amyloid HR mean?

Say you had a variable $X$ that is going up for everyone over time, and that true
risk depends on the *current* value of $X$. 
In an entry time analysis, in a risk set at 10 years all the baseline $X$ values
are out of date, but perhaps they are still approximately correctly ordered.

The COX PL term can always be recentered
$$
	\frac{e^{\beta x_i}}{\sum_j e^{\beta x_j}} = 	
		\frac{e^{\beta (x_i -c)}}{\sum_j e^{\beta (x_j -c)}}
$$
so the HR only depends on differences in $X$.
This back of the envelope justification doesn't work on age scale, we really
should use the current values of $X$.

A second point is the splicing problem.  When I create a curve for the expected
future state of someone who is currently 65.
* at age 80, this will involve subjects enrolled at age 70+ (no one has 20 years
of fu)
* The future for a low amyloid subject will use the risk of someone who was
low amyloid when they were enrolled.  The curve will by systematically too
good.
* We inherit the EKM flaws without noticing.

Poisson approx
* divide age into 5 year bins
* within each bin compute rates as a function of sex, APOE, amyloid, CMC
* for someone age 60 with low amyloid, compute the future state probs
* take a weighted average of hazards at each age

To do
* multi-state model
* additive log hazards?

